%*******************************************************
% Acknowledgments
%*******************************************************
\pdfbookmark[1]{Acknowledgments}{acknowledgments}

\bigskip

\begingroup
\let\clearpage\relax
\let\cleardoublepage\relax
\let\cleardoublepage\relax
\chapter*{Acknowledgments}

Above all, I am most grateful to the LORD, my rock, my fortress, and my deliverer. He is my King from of old. Palpable was His steadfast love, grace, and direction throughout my research.

I am greatly indebted to my supervisors, Dr Deepak Padmanabhan and Dr Anna Jurek-Loughrey, for their innumerable contributions and insights, patience, encouragement, supervision and intellectual support.

I express my gratitude to Dr Paul Miller and Dr Barry Devereux, the latter of whom served as the internal viva examiner, for providing me with constructive criticism and valuable recommendations during my fair share of Annual Progress Review meetings. I am grateful to the external viva examiner, Dr Seun Ajao, for taking the time to examine my thesis, and providing detailed feedback and suggestions, which have enhanced the quality of this work.

I am grateful for the help I have received from members of staff at EEECS. I would like to acknowledge the help of Dr Jesus Martinez Del Rincon and Professor Hans Vandierendonck. Special thanks to Mrs Katie Stewart for her unparalleled support and guidance throughout my time at the school. I am deeply indebted to Mrs Kathleen Ingram for her support, and help with proofreading my work.

In developing this thesis I have benefited from the constructive comments and warm encouragement of friends and colleagues at EEECS. I wish to thank Abdullah, Alimuddin, Ayesha, Maya, Michael, and Pritam. I enjoyed interacting and exchanging ideas with you all.

I am also immensely grateful to friends who have helped and supported me during this journey. Thank you, Abubakar, Akaoma, Chioke, Doreen, Michael, Victor, Victoria, and Udeme. My heartfelt thanks to Kigwab for your warmth and mirth.

Completing this thesis would not have been possible without the great personal support of my family. I am extremely grateful to my parents, who have always nourished and nurtured my curiosities, and to my siblings, for your profound belief in my abilities. You have contributed much more to my work than you realise.

\begin{flushright}
  Martins Dogo \\
  February 2023
\end{flushright}

\endgroup
